A* with Perfect Potentials

### 2020-04-20 paper/ch_potentials.tex
forgot reporting preprocessing
modified deg 2 and deg 3 sections

\usepackage[algo2e,vlined]{algorithm2e}

@@

We observe, that a potential evaluation is always paired with a queue push operation.
By avoiding push operations, we can therefore also avoid potential evaluations.
In this section, we slightly modify how A* works internally.
Our objective is process chains of low degree nodes sequentially without putting them into the queue.

We denote the tentative distance array used in the A* algorithm by $D$.
When removing a node $v_1$ from the A* queue, we iterate over all outgoing edges $(v_1,v_2)$ of $v_1$ and process the chain started by $(v_1, v_2)$.
If $D[v_2] < D[v_1]+w_q(v_1,v_2)$, our algorithm stops processing the chain and continues with the next outgoing edge of $v_1$.
Otherwise, $D[v_2]$ is set to $D[v_1]+w_q(v_1,v_2)$ and processing the chain continues.
If $v_2$ has only one outgoing edge that leads to a node $v_3$ different from $v_1$, we process $v_3$ analogously.
This is repeated until a $v_k$ is reached such that either $D[v_{k-1}]+w_q(v_{k-1},v_k)\ge D[v_k]$ or $v_k$ has more than one successor different from $v_{k-1}$.
Our algorithm continues removing nodes from the queue until a node $x$ is removed for which $D[x]\ge D[t]$.

Our optimization can be described as node contraction performed lazily at query time.
If $G$ is undirected, all nodes with a degree two in $G$ except $s$ are lazily contracted.
As we only employ contraction during the query, we do not loose flexibility.
This contrasts with most hierarchical approaches, that use contraction in the preprocessing step.

Not pushing degree two nodes into the queue gives a moderate to large speedup depending on how many degree two nodes are part of the graph.
We can achieve a slightly larger speedup by also not pushing every degree three node into the queue.

@@

In the previous section, we described how a lazy degree two node contraction can be used to accelerate an A* search.
Denote by $I$ the set of nodes that can be pushed into the A* queue.
Using the $I$ from the previous section, $G\setminus I$ is a disjoint set of paths, if $G$ is undirected.
For performance reasons, it is important that $G\setminus I$ results in a disjoint collection of simple graphs.
Fortunately, we can allow slightly more complex graphs than just paths.
In this section, we use stars where the center node has degree three.
This allows us to further increase the size of $I$, which in turn reduces the number of A* queue operations and potential evaluations.

Let $v_k$ be the last node of a chain as in the previous section.
If $v_k$ has two successor nodes $u$ and $w$ that differ from $v_{k-1}$ and $v_k$ is not part of the queue, then we do not push $v_k$.
Instead, we follow the chains started by $(v_k, u)$ and $(v_k, w)$ as described in the previous section.

@@

\item Applications
	\begin{itemize}
  \item Table
  	\begin{itemize}
    \item columns: running time, potential evals, relaxed arcs, shortest path distance / lower bound, dijkstra running time?
    \item rows: OSM Ger Mapbox Live, forbidden highways, forbidden tunnels, TD Ger06, TD Eur17, DIMACs Eur 1000 {Random|Fastest} Arcs {x2, x10}
		\end{itemize}
	\end{itemize}
\end{itemize}

>>>>

\usepackage{wrapfig}
\usepackage[algo2e,vlined]{algorithm2e}

@@

Potential evaluation is an expensive operation that we want to avoid.
A potential evaluation is always paired with a queue push operation.
By avoiding push operations, we can therefore avoid potential evaluations.
This is the objective of this and the next section.

We modify how A* by processing low degree nodes consecutively without pushing them into the queue.
Our algorithm makes use of the undirected degree $d(x)$ of a node $x$. 
Formally, $d(x)$ is the number of nodes $y$ such that $(x,y)\in E$ or $(y,x)\in E$.

Analogous to A*, our algorithm stores for every node $x$ a tentative distance $D[x]$.
Additionally, it maintains a minimum priority queue. 
Diverging from A*, not all nodes can be pushed but every node has a tentative distance.

Our algorithm differs from A* when removing a node $x$ from the queue.
A* iterates over the outgoing arcs $(x,y)$ of $x$ and tries to reduce $D[y]$ by relaxing $(x,y)$.
If A* succeeds, $y$'s weight in the queue is set to $D[y]+\pi(y)$.
Our algorithm, however, behaves differently, if $d(y)\le 2$.
Our algorithm determines the longest degree two chain of nodes $x,y_1,\ldots, y_k, z$ such that $d(y_i)=2$ and $d(z)\neq z$.
If our algorithm succeeds in reducing $D[y_1]$, it does not push $y_1$ into the queue.
Instead, it iteratively tries to reduce all $D[y_i]$.
If it does not reach $z$, then only $D$ is modified but no queue action is performed.
If $D[z]$ is modified and $d(z)>2$, $z$'s weight in the queue is set to $D[y]+\pi(y)$.

As the target node $t$ might have degree two, our algorithm cannot rely on stopping, when $t$ is removed from the queue.
Instead, our algorithm stops as soon as $D[t]$ is less than the minimum weight in the queue.

@@

\begin{wrapfigure}{r}{3.5cm}

\begin{center}
\includegraphics{push-deg3}
\end{center}

\caption{$x$ is removed from the queue. $y_i$, $z$, $a_i$, and $\alpha_i$ are never part of the queue. $\beta$ and $b$ are pushed into the queue.}
\label{fig:push-deg3}

\end{wrapfigure}

In the previous section, we described an optimized A* variant that does not push degree two nodes.
In this section, we also avoid degree three nodes.

Denote by $x,y_1,\ldots, y_k, z$ a degree chain as described in the previous section.
If $z$ is already in the queue, then $z$'s weight in the queue is adjusted as in the previous section.

If $d(z)\neq 3$ or $z$ is in the queue, our algorithm proceeds as in the previous section.
Otherwise, there exist up to two degree chains $z,a_1,\ldots,a_p,b$ and $z,\alpha_1,\ldots,\alpha_q,\beta$ such that $a_1\neq y_k \neq \alpha_1$.
Our algorithm iteratively tries to reduce all $D[a_i]$ and $D[\alpha_i]$.
If it reaches $\beta$, $\beta$'s weight in the queue is set to $D[\beta]+\pi(\beta)$.
Analogously, if $b$ is reached, $b$'s weight is set to $D[b]+\pi(b)$.
If $b$ respectively $\beta$ are not reached, our algorithm does nothing.
The situation is depicted in Figure~\ref{fig:push-deg3}.

@@

\item Applications
	\begin{itemize}
  \item Table
  	\begin{itemize}
    \item columns: preprocessing time, running time, potential evals, relaxed arcs, shortest path distance / lower bound, dijkstra running time?
    \item rows: OSM Ger Mapbox Live, forbidden highways, forbidden tunnels, TD Ger06, TD Eur17, DIMACs Eur 1000 {Random|Fastest} Arcs {x2, x10}
		\end{itemize}
	\end{itemize}
\end{itemize}
